---
title: "RasterData"
author: "Gonzalo Albaladejo-Robles"
date: "2025-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=F,include=F}
rm(list=ls())
setwd(dirname(rstudioapi::getSourceEditorContext()$path)) # Set the working directory to the directory in which the code is stored
td <- tempdir()
```

Preparation of the environmental information and raster data for the California or Eastern Kingsnake 

* 0. load the needed libraries
```{r echo=T,include=F}
list.of.packages<-c("tidyverse","rstudioapi","data.table","terra","sf","viridis","geodata")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

lapply(list.of.packages,require,character.only=TRUE)
rm(list.of.packages,new.packages)
```

* Load the needed functions
```{r echo=T,include=F}
functions<-"./Functions" %>% list.files(recursive = FALSE,pattern = ".R$",full.names = TRUE)
lapply(functions,function(x) source(x)) #; lapply(functions,function(x)readLines(x) %>% length()) %>% unlist() %>% sum()
```

## 0. Some basic information for the study 
```{r}
crs_p <- "EPSG:4326"
```

## 1. Building a raster----
Raster data usually takes the shape of numerical or categorical matrices that contain spatial references that allow them to characterize a spatial area.

* a. Let's create a raster
First, we create some fake data to populate our matrix. Since we want to populate a matrix with this data we need to make sure that we have enough information for each row and column
```{r}
  r_dat <- rnorm(mean=0,sd=3,n=c(1000*1000)) 
  r_dat <- matrix(r_dat,ncol=1000,nrow=1000)
  
  r_dat <- terra::rast(r_dat)
  r_dat
```

* This is what our raster looks like!
```{r}  
  par(mfrow=c(1,1))
  r_dat %>% plot(col=viridis::inferno(n=200))
```

* b. We have a raster representing some random data, however there is no spatial information stored in it. We need to add more information for it to be useful. Let us try to coerce this raster into our study region
```{r}  
# 1. Load some spatial information
Global <- st_read("./Data/Process_data/General/Glob_map.shp")
Sp_range <- st_read("./Data/Process_data/General/LampCaliforniaeRang.shp")

study_t <- st_read("./Data/Other_ranges/Pro_BaseLayer.shp")
study_t <- study_t %>% st_transform(crs=crs_p)
```  

We can use the vector information to format our raster data
```{r}
  # 2. Coerce our raster to two different spatial extends  
  glob_rast <- r_dat
  # range_rast <- r_dat
  
  ext(glob_rast) <- ext(Global)
  # ext(range_rast) <- ext(Sp_range)
```  
## 3. Raster formatting 

The shape of our raster has changed according to the spatial dimension of our vector data. Not only that, we have changed the data stored in our raster, the extent and resolution fields are now different. For the extent and resolution to make sense, we need to project this data into space (we need to add a Coordinate Reference System)
```{r}
  glob_rast
  crs(glob_rast) <- crs_p
  glob_rast <- terra::project(glob_rast,y=crs_p)
  glob_rast
```  

* a. Check the raster
```{r}
  par(mfrow=c(2,1))
  glob_rast %>% plot(col=viridis::inferno(n=200)) # Now our raster is compatible with the rest of spatial data
  Global %>% st_geometry() %>% plot(add=T,border="white",lwd=1.5) # see how everything line up
  mtext(side=3,adj=0,"a. Non-masked raster",cex=1.2)
  
  glob_mask <- glob_rast %>% mask(mask=Global %>% vect()) 
  glob_mask %>% plot(col=viridis::inferno(n=200)) # We don't want to have data into the ocean, so we are going to mask it
  mtext(side=3,adj=0,"b. Masked raster",cex=1.2)
```

* b. Rasters can be projected and formatted just like vector data.
```{r}
  crs_list <- list("+proj=fouc","+proj=lagrng","+proj=collg",
                   "epsg:3035","+proj=eck1") # These are some cool global projections
  
  lt <- layout(matrix(c(rep(rep(c(1,2,3),each=2),3),rep(rep(c(4,5),each=3),3)),ncol=6,nrow=6),respect = T)
  # layout.show(lt)
  
  for(i in 1:length(crs_list)){
    plot(glob_mask %>% project(y=crs_list[[i]]),col=viridis::inferno(n=200))
    mtext(paste0(letters[i],")"),side=3,adj=0)
  }  
```

* c. However, if we want to have multiple variables, we need to build a `stack` of raster layers, one for each variable. All layers need to have the same parameters (extension, resolution, and crs)
Let's create some extra "data"
```{r}
  # Var 2 
  r_2 <- rast(vals=rnorm(mean=4,sd=0.2,n=c(1000*1000))%>%order(), extent = ext(glob_mask),
                res=res(glob_mask),crs=crs(glob_mask)) %>% mask(mask=Global %>% vect())
  
  # Var 3
  r_3 <- rast(vals=0,res=res(glob_mask),crs=crs(glob_mask),extent = ext(glob_mask)) %>% 
                mask(mask=Global %>% vect(),inverse=T)
  
```

* Combine all the variables into a stack and display the information
```{r}
cool_stack<-c(glob_mask,r_2,r_3)
plot(cool_stack)  
```

## 4. Let us work with some real data

We are going to use the geodata library to download some bioclimatic information from (world-clim)[https://www.worldclim.org/]

* a. Download the data
```{r}
  bio_route <- "./Data/BioClim" ; bio_route %>% dir.create(showWarnings = F,recursive = T)

  bioClim <- geodata::worldclim_global(var="bio",res=5,path=bio_route) # Download the global environmental information
```  

* b. Each dataset consists of 19 different climatic variables with ecological potential to define species niches
```{r}
  names(bioClim)
  bioClim %>% plot()
```  

* c. We don't need the full extension of the dataset, let's crop the raster. We can crop the raster stack using a vectorial object. In this case, we are going to take the data from the united states and mexico
```{r}
# Get the country boundaries using the geodata package
  countries <- geodata::world(resolution=3,level=0,path="./Data/Global")
  countries <- countries %>% st_as_sf()

  countries <- countries %>% filter(GID_0 %in% c("USA","MEX"))  
```

We need to do some data cleaning
```{r}
# The USA is made up of several different polygons, but we are only interested in the larger portion
  USA_split <- countries[2,] %>% st_cast("POLYGON")
  area_USA <- USA_split %>% st_area()
  
  countries[2,] <- USA_split[area_USA==max(area_USA),] #%>% st_geometry() %>% plot()
  # countries %>% st_geometry() %>% plot()
```

Crop the environmental data
```{r}
  c_list <- list()
  for(i in 1:nrow(countries)){
      c_list[[i]]<- bioClim %>% crop(y=countries[i,] %>% vect())
      names(c_list)[i]<-countries[i,"NAME_0"] %>% st_drop_geometry()
      }
```

Visualize the raster tiles
```{r}
  par(mfrow=c(2,1))
  c_list$`United States`$wc2.1_5m_bio_1 %>% plot(col=viridis::viridis(n=200)) ; mtext(side=3,adj=0,"b. United States of America")
  plot(Sp_range %>% st_geometry(),add=T,lwd=2)
  
  c_list$Mexico$wc2.1_5m_bio_1 %>% plot(col=viridis::viridis(n=200)) ; mtext(side=3,adj=0,"a. Mexico")
  plot(Sp_range %>% st_geometry(),add=T,lwd=2)
```

* d. To obtain a full picture of the study area we need to combine both tiles/layers   
Merge crop and plot
```{r} 
 full_rast <- terra::merge(c_list[[1]],c_list[[2]],na.rm=TRUE)
 
  # Crop and plot
  par(mfrow=c(1,2))
  plot(full_rast$wc2.1_5m_bio_1,col=viridis::viridis(n=200))
  plot(Sp_range %>% st_geometry(),add=T,lwd=2)
  mtext(side=3,adj=0,"a. Combined tiles")
```

* e. Crop the raster to the size of the study area
```{r}
  crop_rast <- full_rast %>% crop(y=Sp_range %>% vect())
  plot(crop_rast$wc2.1_5m_bio_1,col=viridis::viridis(n=200))  
  plot(Sp_range %>% st_geometry(),add=T,lwd=2)
  mtext(side=3,adj=0,"a. Study area")
```

With this code, we have downloaded, unified, and adapted several environmental variables, in the form of raster layers. Additionally, we adapted the size and extent of this data to our study area.

## 5. Extract data from a raster file

Raster data behaves like matrix data in R. Therefore, we can extract values using their column and row coordinates (e.g. `valueij=raster[i,j]`). If we apply this to a raster stack or spatStack object we get the value of `ij` cell for the "n" number of variables/layers included in that object. Although this is useful in some cases, the spatial nature of the raster format allows us to extract this information using other spatial objects such as points, polygons, lines, etc. 

* a. Load some point data
We are going to load the distribution points of L. californiae in its natural range
```{r}
  # a. Load some point data
  Sp_points <- st_read("./Data/Process_data/Records/Gbif_rec.gpkg")  
```  
  
* b. extract the information from the raster
```{r}
  Sp_data <- terra::extract(crop_rast,Sp_points %>% vect())
  head(Sp_data)
```  

# 6. Operations with rasters

Same as matrix objects, we can perform mathematical and statistical operations over raster data. These calculations can be made individually (one layer at a time) or across layers, and also using the whole layer to produce individual numerical results (e.g `mean(raster)`, a global mean is calculated for the whole raster) or across cells to produce a new spatial raster (e.g `log1p(raster)`, raster cell values are transformed and a raster is returned).

* a. As an example, we are going to calculate the NDVI for the invasive range of L. californiae using some satellite data

load the satellite data
```{r}
Sentinel_routes <- "./Data/Sentinel" %>% list.files(pattern=".tif$",recursive = TRUE,full.names = TRUE)
  sentinel_rast <- lapply(Sentinel_routes,rast)
```  

Check the data
```{r}

```

* b. Combine the rasters and harmonize them with the rest of the spatial information
```{r}
  strip1 <- merge(sentinel_rast[[1]],sentinel_rast[[2]]) 
  strip1 <- strip1 %>% project(y=crs_p) %>% crop(y=study_t %>% filter(Island=="Gran_Canaria") %>% st_geometry())
  
  strip1 <- strip1 %>% mask(mask=study_t %>% filter(Island=="Gran_Canaria") %>% vect())
  strip1 %>% plot(col=viridis::turbo(n=200))
```

* c. Calculate the NDVI using the formula NDVI = (B8-B4)/(B8+B4)
```{r}
  NDVI_strip1 <- (strip1$`B08-1`-strip1$`B04-1`)/(sum(strip1)) ; names(NDVI_strip1) <- "NDVI"
```

Check the results
```{r}
  plot(NDVI_strip1,col=viridis(1000),box="n",axes=F,main="NDVI Gran Canaria")
```  
