---
title: "Vector Data"
author: "Gonzalo Albaladejo-Robles"
date: "2025-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list=ls())
setwd(dirname(rstudioapi::getSourceEditorContext()$path)) # Set the working directory to the directory in which the code is stored
td <- tempdir()
```
# Preparation of the record data and spatial information for the SDM of Lampropeltis californiae   

* 0. load the needed libraries
```{r}
list.of.packages<-c("tidyverse","rstudioapi","data.table","pdftools","sf")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

lapply(list.of.packages,require,character.only=TRUE)
rm(list.of.packages,new.packages)
```

* 0.1 Load the needed functions
```{r}
functions<-"./Functions" %>% list.files(recursive = FALSE,pattern = ".R$",full.names = TRUE)
lapply(functions,function(x) source(x)) 
```

We are going to use plannar calculations along the whole code. This may cause a lot of warnings to pop up, but it simplifies some calculations and ensures the correct functioning of several functions.
```{r}
sf::sf_use_s2(FALSE)
```

# 0. Load some general spatial information
We are going to use the `geodata` package to download some public spatial information directly into `R`. In this case we are going to download the general shape of all countries boundaries

* a. Download the spatial layout of the world land-masses
```{r}
"./Data/Global" %>% dir.create(recursive = T,showWarnings = T)
global <- geodata::world(resolution = 1,level=0,path="./Data/Global")
```

The global data is a `spatvector` object. This class of spatial object is compatible with the `terra` package which is mainly oriented to spatial raster generation and processing. Althought `spatvector` objects offer a lot of functionallity, we are going to transform it into an `sf` object. Both `sf` and `terra` packages are part of a major efford to consolidate and standarize spatial data managing and processing in `R`. These two libraries offer a faster a more natural approach to spatial objects and substitude previous libraries such as `raster`, `maptools`, `rgdal`, and `rgeos`. Additionally, both `sf` and `terra` are computationally faster and easy to combine

Tranform the `spatVector` into an `sf` object
```{r}
global <- st_as_sf(global)
global
```

The data is displayed as a `data.frame`.

* b. Check the information
```{r}
plot(global %>% st_geometry(),col="grey5") # We get an error!
global <- global %>% st_make_valid() %>% st_union() # there are other tricks that sometimes work
```

```{r}
plot(global %>% st_geometry(),col="grey5")
g = st_graticule(global) ; plot(g[1], add = TRUE, col = 'grey') # add a spatial grid
```

* c. Spatial data need to be projected or referenced in order to be useful. Different projections serve different purposes
```{r}
  crs_list <- list("+proj=fouc","+proj=lagrng","+proj=collg",
                   "+proj=chamb +lat_1=10 +lon_1=30 +lon_2=40","+proj=eck1") # These are some cool global projections
  
lt <- layout(matrix(c(rep(rep(c(1,2,3),each=2),3),rep(rep(c(4,5),each=3),3)),ncol=6,nrow=6),respect = T)
  # layout.show(lt)
  par(par=c(0,0,0,0))
  for(i in 1:length(crs_list)){
    plot(global %>% st_transform(crs=crs_list[[i]]) %>% st_geometry(),col="gold3",border=NA)
              g = st_graticule(global %>% st_transform(crs=crs_list[[i]]))
              plot(g[1], add = TRUE, col = 'grey25',xpd=T)
              mtext(paste0(letters[i],")"),side=3,adj=0)
              }  
```

This global data is going to help us put our data into context globally  
      
# 1. Lampropeltis local data
We have some data from a project that aims to control an alien invasive species that was introduced into an island:

* a. The data is presented as a csv file with 4 variables (type of records, date, X coordinate, and Y coordiante)
```{r}
data <- read.csv(paste("./Data/Life+ Lampropeltis","Raw_records.csv",sep="/"))
    head(data)
```

* b. We need to transform this CSV into a projected spatial object that can interact with the rest of spatial data
```{r}
    crs_life <- "EPSG:4326" # This data is formatted using the UTM reg-can 95 EPSG:4326
    PointsLife <- data %>% st_as_sf(coords=c("Coorde.nadaX","coorden.adaY"),crs=crs_life) 
```

* c. Now that we have the points, we need some other spatial information to add context to this information
```{r}
    LifeRANGE <- st_read("./Data/Other_ranges" %>% list.files(pattern=".shp$",full.names = T))
  
  # d. Plot the information
    par(mfrow=c(1,1))
    plot(LifeRANGE %>% st_geometry())
    plot(PointsLife %>% st_geometry(),pch=19,col="tomato3" %>% adjustcolor(alpha.f = 0.10),add=T)
```    

The general map is ok but we can add more information, such as the name of the islands

* c.1 Get the polygon centroids
```{r}
    centroids <- st_centroid(LifeRANGE) %>% st_coordinates() # To allocate the labels we are going to take the centroids of the polygons as a guide
    LifeRANGE <- cbind(LifeRANGE,centroids) # Lets add this to the polygon information
    head(centroids) # X Y coordinates
```    

* c.2 Add the polygons names to the general map
```{r}
    plot(LifeRANGE %>% st_geometry()) # Now lets use the coordiantes of the centroids to allocate the labels
    
    text(x=LifeRANGE$X,y=LifeRANGE$Y,labels=LifeRANGE$Island,col="blue4") # the location is a bit off, lets add some space in the Y axis
    text(x=LifeRANGE$X,y=LifeRANGE$Y+40000,labels=LifeRANGE$Island,col="red",cex=0.75,xpd=T)
```  

* c.3 The map is way to general and offer litle detail about the distribution of the species. Lets subset the spatial data so we only display the polygon in which presence data is present
```{r}
    plot(LifeRANGE %>% filter(Island=="Gran_Canaria") %>% st_geometry()) # subset the spatial information
    plot(PointsLife %>% st_geometry(),pch=19,col="tomato3" %>% adjustcolor(alpha.f = 0.10),add=T)
```

* c.4 Put everything together in a fancy way
```{r}
    tl<-layout(matrix(c(2,rep(1,times=8)),ncol = 3,nrow = 3,byrow = T))
    layout.show(tl)
    par(bg="white")
    
    # Study area map
    plot(LifeRANGE %>% filter(Island=="Gran_Canaria") %>% st_geometry()) # subset the spatial information
    plot(PointsLife %>% st_geometry(),pch=19,col="tomato3" %>% adjustcolor(alpha.f = 0.10),add=T) # Add the labels
    mtext(side=3,adj=1,"b. Gran Canaria Island") # Panel legend
    
    # Archipelago map
    par(bg=NA)
    plot(LifeRANGE %>% st_geometry(),col=ifelse(LifeRANGE$Island=="Gran_Canaria","tomato3","grey88"))
    text(x=LifeRANGE$X,y=LifeRANGE$Y+40000,labels=LifeRANGE$Island,col="black",cex=0.75,xpd=T) # Add the labels
    mtext(side=3,adj=0,"a. Canary Islands") # Panel legend
    par(bg="white")
```

Plots are a fundamental part of dealing with spatial information. Same as with any other analysis or data-processing exercise, data visualization can help us to detect patterns, errors, or unknown relationships. We are going to do a lot of plots throught this code

# 2. Load the spatial data from the IUCN Red-List
Portals such as the IUCN Red List of Threatened species contain a open access, useful, and curated spatial data. In this case we downloaded all the range maps belonging to reptiles (squamata). This and other information can be found at https://www.iucnredlist.org/resources/spatial-data-download

* a. Unzip and load the IUCN spatial data
```{r}
  pol_exit <- paste(td,"IUCN_pols",sep="/") # Create a temporal folder to host the data
  dir.create(pol_exit,recursive = TRUE,showWarnings = FALSE)

  # Lets unzip the REPTILE spatial data from the IUCN Red-List
  "./Data/IUCN_ranges" %>% list.files(pattern=".zip$",full.names = TRUE) %>% unzip(exdir=pol_exit)
  route_IUCN <- pol_exit %>% list.files(pattern=".shp$",full.names = TRUE)
```

* a.1 Read and combine 
```{r}
  IUCN_ranges <- rbind(route_IUCN[1] %>% st_read(),route_IUCN[2] %>% st_read())
  IUCN_Lamp <- IUCN_ranges %>% filter(genus=="Lampropeltis")
```

The dataset contains spatial information for ``{r} length(IUCN_Lamp$sci_name %>% unique())`` species of reptiles distributed across the globe. However, for this exercise we are only interested in Lampropeltis californiae so we are to retrieve the spatial information for this species and its close relatives (Genus level).
```{r}
  SpeciesIUCN <- IUCN_Lamp$sci_name %>% unique()
```

# 3. Get the GBIF records (Skip this for the seminar!)
Now that we have the list of species belonging to the Lampropeltis genus for which we have spatial range data, we are going to download the presence records of these species from GBIF (ref here).

GBIF (Global Biodiversity Information Facility)is an international network and data infrastructure aimed at providing open access data about biodiversity. This portal contains taxonomic as well as distribution information for millions of species across the planet.

* a. Using the `rgbif` package we can connect to GBIF servers and download the spatial records of the species we are interested in. But first we need to check the taxonomy of our species and retrieve all the possible synonyms
```{r}
  crs_gbif <- "EPSG:4326"
  
  # a. Get he GBIF spatial records for our species ----
  "./Data/Gbif" %>% dir.create(recursive=T,showWarnings = F)
  
  # b. Get the taxonomic information for the species ----
  TaxInformation <- lapply(SpeciesIUCN,function(x) retrieve_syns(x,n_times=50))
  TaxData <- lapply(TaxInformation,function(x) x[["TaxDat"]]) ;  TaxData <- do.call("rbind",TaxData)
  
    names_sp <- list()
    for(i in 1:nrow(TaxData)){
      # Combine the names
      names_l <- TaxData[i,colnames(TaxData) %in% c("Or_name","IUCN_name","IUCN_syn","ITIS_name","ITIS_syn")] %>% paste(collapse=";")
      
      # Remove the NAs and create a list of names
      names_l <- strsplit(names_l,split=";") %>% unlist()
      
      # Remove the NAs and duplicated names
      names_l <- names_l[!names_l %in% "NA"]
      names_l <- unique(names_l)
      
      names_sp[[i]] <- list(Or_name=TaxData[i,"Or_name"],All_names=names_l)
      
      }
```

* b. Export the taxonomic information for the different species
```{r}
  
    "./Data/Sp_info" %>% dir.create(recursive = TRUE,showWarnings = FALSE)
    
    write_rds(names_sp,file=paste("./Data/Sp_info","Species_names.rds",sep="/"))
    write.csv(TaxData,paste("./Data/Sp_info","Species_tax.csv",sep="/"))
```

* c. With the species information, download the GBIF spatial records
```{r}
  
    for(i in 1:length(names_sp)){
      print(paste("Getting the information for",names_sp[[i]][["Or_name"]]))
      try(Dowload_gbif(sp_list = names_sp[[i]][["All_names"]],
                       exit_route = "./Data/Gbif",
                       initial_date = 1970),
          silent=F)
    }
```  

* c.1 Load and unified the spatial and taxonomic information for the species (jump here directly) ----
```{r}
    crs_gbif <- "EPSG:4326"
    GBIFrecords <- lapply("./Data/Gbif" %>% list.files(pattern=".csv",full.names = TRUE),read.csv)
    GBIFrecords <- GBIFrecords %>% rbindlist(fill=T)
  
    GBIFrecords <- st_as_sf(GBIFrecords,coords=c("decimalLongitude","decimalLatitude"),crs=crs_gbif)
```

* d. Lets have a look a the GBIF data
```{r}
  # d. Display all the records within the boundaries of the species ranges----
    # This is nice, but we are only interested in the area covered by the range of our species
    crop_ranges <- global %>% st_crop(st_bbox(IUCN_Lamp)) # We are going to use the bounding box of the range data to     crop the global boundaries data
    
    # Create a color palette for the species ranges
    pal_species <- data.frame(species = unique(IUCN_Lamp$sci_name),sci_name = unique(IUCN_Lamp$sci_name),
                              col = viridis::turbo(n=length(unique(IUCN_Lamp$sci_name))))
    
    pal_species[pal_species$species=="Lampropeltis californiae","col"] <- "tomato3"
    
    IUCN_Lamp <- merge(IUCN_Lamp,pal_species,by="sci_name")
    GBIFrecords <- merge(GBIFrecords,pal_species,by="species")
    
  # d.2 Build the plot ----
  # d.2.a Global map ----
    # General polygon
    par(mfrow=c(1,1))
    plot(crop_ranges %>% st_geometry())
    
    # Polygon ranges
    plot(IUCN_Lamp %>% filter(sci_name == "Lampropeltis californiae") %>% st_geometry(),
         col="tomato3" %>% st_drop_geometry() %>% adjustcolor(alpha.f = 0.75),key.pos=2,
         cex.axis=0.5,key.length=0.4,add=T)
    
    plot(IUCN_Lamp["sci_name"] %>% filter(sci_name != "Lampropeltis californiae"),
         col=IUCN_Lamp$col[IUCN_Lamp$col!="tomato3"]%>%adjustcolor(alpha.f = 0.25),add=T)
    
    # Add legend to the plot
    legend("bottomleft",legend=pal_species$species,
           col=pal_species$col %>% adjustcolor(alpha.f = 0.75),
           pch=15,cex=ifelse(pal_species$species=="Lampropeltis californiae",0.75,0.5),ncol=2,
           pt.cex = 1.2,title="Species",bty="n")
```

* d.2.b Species map

As we can see from the general map, the range of Lampropeltis californiae overlaps with other species of the same genus. This can lead to identification and classification errors
```{r}
    Over_ranges <- IUCN_Lamp %>% filter(sci_name!="Lampropeltis californiae")
    Sp_range <- IUCN_Lamp %>% filter(sci_name=="Lampropeltis californiae")
    
    intersect_pols <- st_intersects(Sp_range,Over_ranges,sparse = T)
    Over_ranges <- Over_ranges[intersect_pols[[1]],]
```

Check the spatial distribution
```{r}
    # d.2.b.1 Check the spatial distribution ----
    Sp_range %>% st_geometry() %>% plot(col=Sp_range$col)
    Over_ranges %>% st_geometry() %>% plot(col=Over_ranges$col %>% adjustcolor(alpha.f = 0.5),add=T)
    plot(crop_ranges %>% st_geometry(),add=T)
    
    study_area <- rbind(Sp_range,Over_ranges)
    
    #
    legend("bottomleft",legend=study_area$sci_name %>% unique(),
           col=study_area$col %>% unique() %>% adjustcolor(alpha.f = 0.5),
           pch=15,cex=ifelse(study_area$sci_name=="Lampropeltis californiae",0.85,0.5),ncol=2,
           pt.cex = 1.2,title="Species",bty="n")
```

# 4. Clean the GBIF records

* a. Which distribution points overlap with the species range?
```{r}
  records_range <- GBIFrecords[st_intersects(GBIFrecords,Sp_range,sparse = F),]
  all_recs <- xtabs(~records_range$species) # Number of records for each species within the species range
  sum(all_recs)
```

As the data shows, we have an overlap of several close related species within the range of Lampropeltis californiae. This can cause erroneous identification of individuals and therefore erroneous point records. Therefore we need to filter the spatial records to try to avoid these issues:
  
* b. Which portions of the species range can be inhabited by a close related species?
```{r}
sub_ranges <- st_intersection(Over_ranges,Sp_range) # Crop the portions of the Species ranges that overlap with L. californiae
  
  mInt <- st_intersects(records_range,sub_ranges,sparse = F) # Which points overlap with the other species ranges
  head(mInt) # We have a column of indexes for each polygon/ sub-polygon
  mInt <- apply(mInt ,1,function(x) as.numeric(x)%>% sum()) # We are going to combine all the indexes into a single one
  mInt <- ifelse(mInt >= 1,TRUE,FALSE)
  sum(mInt) # Total number of observations withint the ranges
```  

* c. However, since we don't trust the ability of herpetologist to correctly classified close-related species, we are going to add a buffer around the overlapping species ranges to exclude the points that fall into these areas.
```{r}  
  sub_ranges_buff <- sub_ranges %>% st_buffer(dist=0.15) %>% st_intersection(Sp_range) # Add a buffer around the overlapping areas to remove problematic records
 
  # Which distribution points are not in conflict areas (overlapping with other species)?---- 
  mInt <- st_intersects(records_range,sub_ranges_buff,sparse = F)
  mInt <- apply(mInt ,1,function(x) as.numeric(x)%>% sum()) # We are going to combine all the indexes into a single one
  mInt <- ifelse(mInt >= 1,TRUE,FALSE)
  sum(mInt)
```

filtering the observations
```{r}
  nrow(records_range)-(sum(mInt)) # Number of records within the range of the species and out of conflict areas
  clean_records <- records_range[!mInt,]
```

Check the results
```{r}
  # par(mfrow=c(1,2))
  lt<-layout(matrix(c(1,1,2,3),ncol=2,nrow=2),respect = T)
  # layout.show(lt)
  
  plot(Sp_range %>% st_geometry(),col="tomato3") ; mtext(side=3,adj=0,"a) Range of Lampropeltis californiae",cex=0.85)
  plot(sub_ranges %>% st_geometry(),col=sub_ranges$col %>% adjustcolor(alpha.f = 0.75),add=T)
  plot(sub_ranges_buff %>% st_geometry(),col=NA,border=sub_ranges$col %>% adjustcolor(alpha.f = 0.85),lty=3,add=T)
  
  legend("bottomleft",legend=sub_ranges$sci_name %>% unique(),pch=15,
         col=sub_ranges$col %>% unique() %>% adjustcolor(alpha.f = 0.75),
         title="Species Ranges",bty="n",pt.cex=1.5,cex=0.7)
  
  legend("left",legend="Buffered area",lty=3, col="tomato3" %>% adjustcolor(alpha.f = 1),bty="n",lwd=1.5,cex=0.7)

  # Plot the clean points
  plot(Sp_range %>% st_geometry(),col="grey70") ; mtext(side=3,adj=0,"b) `Clean records`",cex=0.85)
  plot(sub_ranges %>% st_union() %>% st_geometry(),border=NA,col="white" %>% adjustcolor(alpha.f = 0.65),add=T)
  
  plot(clean_records %>% st_geometry(),col=clean_records$col,add=T,pch=19,cex=0.5)
  
  # Plot the discarded points
  plot(Sp_range %>% st_geometry(),col="grey70") ; mtext(side=3,adj=0,"c) Discarded records",cex=0.85)
  plot(sub_ranges %>% st_union() %>% st_geometry(),border=NA,col="white" %>% adjustcolor(alpha.f = 0.65),add=T)
  
  plot(records_range[mInt,] %>% st_geometry(),col="black",add=T,pch=3,cex=0.2)
```  

* d. Now we can trust herpetologists on correctly classifying our snake species
```{r}
  clean_records <- clean_records %>% filter(species=="Lampropeltis californiae")
  nrow(clean_records)
```


# 5. Unified the spatial data

Now that we have our GBIF data and the specific data from our study, we need to make them compatible. This is adjusting their general format and the specific spatial parameters that define their spatial projection

* a. Check the spatial objects
```{r}
  LifeRANGE %>% st_crs() == clean_records %>% st_crs() # CRS are different 
  
  LifeRANGE %>% st_crs()
  clean_records %>% st_crs()
```

* b. Transform the spatial information to the same CRS
```{r}
  LifeRANGE <- LifeRANGE %>% st_transform(crs=crs_gbif) # We are going to use the CRS from the GBIf data
```

* c. We want a general polygon that includes both the native range of the species and the introduced area
```{r}
  full_range <- c(IUCN_Lamp %>% st_geometry(),LifeRANGE %>% st_geometry())
  glob_lamp <- global %>% st_crop(st_bbox(full_range))      
```

* d. Lets put all together
```{r}
  par(mfrow=c(1,1))
  
  glob_lamp %>% st_geometry() %>% plot()
  Sp_range %>% st_geometry() %>% plot(col="tomato3",add=T)
  LifeRANGE %>% filter(Island=="Gran_Canaria") %>% st_geometry() %>% plot(col="tomato3",add=T)
  
  # Add some labels
  invasive_point <- LifeRANGE %>% filter(Island=="Gran_Canaria") %>% st_centroid() %>% st_coordinates()
  native_point <- Sp_range %>% st_centroid() %>% st_coordinates()  

  arrows(x0=invasive_point[1],y0=invasive_point[2]+1,x1=invasive_point[1]-5,y1=invasive_point[2]+5,lwd=2,code=0,col="tomato3")
  text(x=invasive_point[1]-5,y=invasive_point[2]+7,"Invasive range",col="tomato3")
  
  arrows(x0=native_point[1],y0=native_point[2]+1,x1=native_point[1]+7,y1=native_point[2]+7,lwd=2,code=0,col="tomato3")
  text(x=native_point[1]+8,y=native_point[2]+9,"Native range",col="tomato3")
   
  # Wouldn't be cool to calculate the distance between the  native and the invasive range?
  distance_sites <- st_distance(Sp_range %>% st_centroid(),
                                LifeRANGE %>% filter(Island=="Gran_Canaria") %>% st_centroid())
  units(distance_sites)<-"km"
  
  arrows(x0=native_point[1],y0=native_point[2],x1=invasive_point[1],
         y1=invasive_point[2],lwd=2,code=0,col="grey10",lty=2)
  
  text(x=mean(c(native_point[1],invasive_point[1])),y=mean(c(native_point[2],invasive_point[2]))+2, 
       label=paste(distance_sites %>% round(digits=2),"km"),srt=-3,font=2)
```

# 6. Export the spatial data

All our spatial information is following the same format and spatial projection
```{r}
  exit_dir <- "./Data/Process_data" ; exit_dir %>% dir.create(recursive = T,showWarnings = F)
  
  # As with any other R object, if we want to preserve the processed spatial objects we need to export/ save them
  # General polygons
  paste(exit_dir,"General",sep="/") %>% dir.create(recursive = T,showWarnings = F)
  
  st_write(IUCN_Lamp,paste(exit_dir,"General",paste0("Lampropeltis_RANGES",".shp"),sep="/"),append=FALSE) # IUCN ranges for Lampropeltis Genus
  st_write(Sp_range,paste(exit_dir,"General",paste0("LampCaliforniaeRang",".shp"),sep="/"),append=FALSE) # Range of L. californiae
  st_write(glob_lamp,paste(exit_dir,"General",paste0("Glob_Dist",".shp"),sep="/"),append=FALSE) # Lampropeltis ranges
  st_write(global,paste(exit_dir,"General",paste0("Glob_map",".shp"),sep="/"),append=FALSE) # General Global polygon
  
  # Clean records
  paste(exit_dir,"Records",sep="/") %>% dir.create(recursive = T,showWarnings = F)
  
  st_write(clean_records,paste(exit_dir,"Records",paste0("Gbif_rec",".gpkg"),sep="/"),append=FALSE) # Gbif Clean Records
  st_write(PointsLife,paste(exit_dir,"Records",paste0("Study_rec",".shp"),sep="/"),append=FALSE) # Study Clean Records
```  

End of the script
  
